From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-01-11 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Update classical-phil-article.tex
Extensive edits to Section 4: rewriting intro, rewriting section 4.2 (incorporating gauge redundancy argument from section 5.2); edits to 4.1 and 5.2.

As indicated at the start of Section~\ref{privileging}, we use ``fundamentality" in a non-metaphysical fashion. On our usage, differences in fundamentality amount to differences in generality.\footnote{Of course, we allow for other non-metaphysical uses of ``fundamental," tracking differences in epistemic or conceptual primacy. For instance, we can think of the canonical position and momentum observables as ``epistemically fundamental" because they suffice to completely specify a classical state, fixing its values for all other observables. \textcites[31, 200]{Ruetsche} refers to this as ``fundamental in the physicist's sense." Section~\ref{density},discusses another sense in which the Hamiltonian formalism is more fundamental than Lagrangian mechanics.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of ``structure." Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires that fewer physical conditions are satisfied. Compared with North's \parencites*[]{North} account of structure, our definition has two distinct advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which empiricists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

* ``Fundamentality" because it makes the physics more prominent. Some people use coordinate-free/independence as necessary conditions for fundemantaly/physical significance. This means coordinates do not appear in the definition. Yet, densities require coordinate for the definition, therefore cannot be coordinate-free/independent, therefore there are some physically significant quantities that do not satisfy that. Why fundamental??? (reverse mathematics? fundamental because it recovers the correct math from the physical assumption?)

@@

\subsection{Density and Riemmanian geometry}

It is helpful to take a quick look at how densities work in general so that we can better understand what is so special about them in the context in Hamiltonian mechanics. By density we mean the intuitive physical notion of a quantity divided by the region it occupies, like mass per volume or charge per area. Over a continuum, this is an integrable function $\rho(x)$. Mathematically, the concept is captured by the Radon–Nikodym derivative: we have a region $U$, we have two measures $m$ and $V$ and the derivative satistifies $m(U) = \int_U \rho dV$. This makes it clear that the units of $\rho$ depends on the units of both $m$ and $V$. More precisely, if $\rho(x, y)$ is a function of multiple variables, we will want $\int_U \rho(x, y) dx dy$ so that the units of $\rho$ are consistent with the units of the arguments. For example, in polar coordinates $\rho(r, \theta)$ will return mass over meter-radians if we measure $r$ in meters and $\theta$ in radians.

The density is coordinate dependent because its definition requires us to make a choice of coordinates. Worse than that, it is not even coordinate invariant: the density changes with the Jacobian. For example, changing from polar to Cartesian coordinates we have $\rho(x,y) = |J| \rho(r, \theta) = r \rho(r, \theta)$. Since the Jacobian is a function of the point, this makes the value of the density, in general, not directly comparable at different points. If we have a uniform density in $x$ and $y$, we will have an $r$-dependence in $r$ and $\theta$. This dependence is introduced precisely because the area identified by $dr$ and $d\theta$ is $dr \,r d\theta$, which is a function of $r$.

In differential geometry, a different mathematical object is called the `density,' namely a volume form (i.e. an alternating function from $n$ vectors to a real number) that is unaffected by changes in orientation. Our physically significant density $\rho$ is actually the component of this volume form $\nu = \rho(x^i) dx^1 \wedge dx^2 \wedge ... \wedge dx^n$. Note that $\nu$ combines the density and the volume into a coordinate-free object. Considering the total mass in the region $U$, we have $m(U) = \int \nu$ so essentially $\nu = dm$ which is the infinitesimal mass contribution we are integrating. But this means the value returned by $\nu$ depends not only on how must stuff there is at a given point, but also on the size of the volume identified by the $n$ vectors given to the form. [Not immediately clear to me what the upshot is of this last point; e.g. how does this relate to $\nu$ being coordinate-free?]

In Riemannian geometry, we have a canonical volume form $dV = \epsilon \sqrt{|g_{ab}|} dx^1 \wedge dx^2 \wedge ... \wedge dx^n$ where $|g_{ab}|$ is the determinant of the metric tensor as expressed in local coordinates, and $\epsilon$ is $1$ or $-1$ depending on the orientation. In component notation, we can express the volume form as $\sqrt{|g_{ab}|}\epsilon_{a_1 ... a_n}$, where $\epsilon_{a_1 ... a_n}$ is the Levi-Civita symbol. We can then express any differential density $\nu = \hat{\rho} dV$ as the product of the volume density times a scalar function $\hat{\rho}$. At first glance, this expression misleadingly suggests that the density $\nu$ is coordinate-independent, but this is not the case. [double check that this claim applies to $\nu$ rather than $\hat{\rho} $. Is this in tension with saying earlier that $\nu$ is coordinate-free?] The quantity $|g_{ab}|$ depends on the choice of coordinates and is, in general, a function of the point. However, we can always find a set of coordinates that, at least locally, makes $|g_{ab}|$ constant. The determinant $|g_{ab}|$ is the Jacobian of the transformation to those coordinates [the rationale/point of this sentence was not clear to me]. For example, in Cartesian coordinates $|g_{ab}| = 1$ and the volume form $dV = dx \wedge dy$; in polar coordinates $|g_{ab}| = r^2$ and the volume form $dV = r dr \wedge d\theta$ [above we said that the Jacobian was $r$ rather than $r^2$]. This means that $\hat{\rho}$ is really expressing the density in a particularly convenient Cartesian-like coordinate system: one in which we can compare the density point-wise. We can also see this if we look at the units of $\hat{\rho}$, which are in terms of inverse volume as expressed in the privileged system.\footnote{This is similar to what happens with the concept of proper time: it is defined in the comoving frame, so it is coordinate dependent. Yet, the expression $d\tau = \frac{1}{c} \sqrt{g_{ab} dx^a dx^b}$ is coordinate invariant.}

We can now look at what happens for distributions over position and velocity. Note that the velocities transform in the same way (i.e. contravariantly) as differentials in position. Therefore $\rho(x^i, v^j) = \sqrt{|g_{ab}|} \sqrt{|g_{ab}|} \rho(\hat{x}^i, \hat{v}^j) = |g_{ab}| \rho(\hat{x}^i, \hat{v}^j)$ $dx^a$. The determinant of the metric tensor is therefore directly the volume of the tangent bundle [I didn't follow this point]. The relationship to the density as expressed in phase space is therefore the following:
\begin{equation}
	\rho(q^i, p_j) = |g_{ab}| \rho(x^i, v^j).
\end{equation}
A change of coordinates will induce a contravariant change in the velocity and a covariant change in the momentum. The left side will remain invariant. The two terms on the right side will change in the opposite way so that their product is invariant. Focusing on densities, then, brings about a physically-motivated way to understand the relationship between the cotangent and the tangent bundle, where the metric tensor plays a role in defining the geometry of the second in a similar way that the symplectic form defines the geometry of the first: it defines the count of states so that densities can be properly tracked.

---

@@

*[Could save the following for discussion of gauge redundancy in section 5.2] Not all ``state variables" are physical. Hamiltonian mechanics does not really care what the q's and p's are. It is the motion assumption that tells us that q's are position, simply because they define the unit system. P is a guage quantity (we care about the trajectory, the velocity). In this view, p is simply a ``label'' we can use to express the density over states. Therefore we do care about the physicality of the invariant distribution more than the physicality of the state variables themselves. $->$ Invariant density is really the important physical element/object in Hamiltonian mechanics.

* Units are physically significant. In fact, the ability to express the densities in units is what gives us the symplectic structure. It's a physical requirement that we must be able to express the density in some unit.

* Invariant densities: physical significance connected to coordinate invariance. [incorporated]

* Density comparison is independent of coordinates. We can compare densities point-wise without worry. [incorporated]

* Makes the fundamental role that phase space plays in statistical mechanics more physically salient/perspicuous. Areas/volumes are measures of number of states. They are coordinate invariant. Also the Shannon/Gibbs entropy are coordinate invariant [but not coordinate independent?] (see physicality of quantities that are coordinate invariant)

* Close the section with a more general discussion on physicality/fundamentality and all those other physical goodness.

We have now substantiated the claim of Section~\ref{privileging} that Hamiltonian mechanics requires less structure than Lagrangian mechanics, in virtue of requiring fewer physical conditions. While this provides one reason for privileging Hamiltonian mechanics, it is not the only one. A further---possibly even more important---reason stems from the relationship between Hamiltonian mechanics and invariant densities. As we argue in this section, viewing physics through the lens of invariant densities has much to commend it. In general, it provides a better understanding of classical systems. Specifically, we will see that densities are of fundamental physical significance due to their coordinate-invariance. Hamiltonian mechanics---unlike Lagrangian mechanics---allows us to compare these densities point-wise in any coordinate system. Although the density is coordinate-invariant, it is not coordinate-independent (it cannot be given a coordinate-free definition). This is because expressing the density requires a choice of units, a fact that is closely connected with symplectic structure. Hence, we will argue that some coordinate-dependent representational choices are nevertheless physically significant. This view is at odds with metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms [get citations: Field 1980; Chen papers; North]. 
% Section~\ref{action} uses the invariant density to provide a geometrical interpretation of the action principle.
% Remember: the density is coordinate-invariant but not coordinate-independent; it is actually coordinate-dependent (i.e. it does not admit of something like a coordinate-free definition)
% Also remember: there are quantities that are both coordinate-invariant and coordinate -dependent that are still physically significant, e.g. choices of q's and p's (?)
% Coordinate-invariant in general means that definition of the quantity involves referencing coordinates, but if you change the coordinates, then the value of the quantity remains the same. (E.g. transforms as a scalar). so densities are never coordinate-independent. defining a density requires a choice of units, so involves referencing units. 
%see 57.3 minutes of 12-21 recording: when we define the density originally, it is coordinate-free (rho as a function of the state space of the particles, S). But when we go to express it numerically, then it becomes coordinate-dependent. (I.e. when we express it over a manifold, then it needs to be a density for integration, and hence it becomes coordinate-dependent). 
% see 51 minutes of 12-21 recording: Densities (the coordinate-dependent kind) are physically significant partly because they tell you how much stuff is in each state

In the Hamiltonian formalism, the density has the following desirable property: we can compare the density point-wise across phase space. If the scalar value of the density at one point is larger than another, this entails that the density is higher at that point of phase space. Physically, this means that there is more physical stuff at this point. Likewise, if the scalar value of the density is the same at two different points, this entails that there is the same amount of stuff at those points. For brevity, will refer to this as the property of \textit{point-wise comparison}. Since in general the density is coordinate-dependent, this property of point-wise comparison does not always hold. For instance, it can fail as we change from Cartesian to polar coordinates [figure out where to insert or reference the illustration of this shown above]. For point-wise comparison to hold, we require a coordinate-invariant density. As the derivation in Section~\ref{deterministic} shows, we must define our state space using a symplectic manifold in order to define an invariant density. In other words, to define a density with the property of point-wise comparison, we must use canonical coordinates $(q, p)$ defined on a symplectic manifold. As we will discuss further in Section~\ref{gauge}, the conjugate momentum $p$ is itself a gauge-dependent quantity. Hence, the real physical significance of canonical coordinates is that they allow us to define an invariant phase space density $\rho $. [Move this section 5.2 stuff up into this section; and figure out how to reorganize this paragraph]

It is instructive to contrast this situation with Lagrangian mechanics. In this case, the density depends on the position and velocity state variables $(q,\dot{q}) $. Crucially, these state variables transform contravariantly, rather than covariantly. Consequently, the density $\rho(q,\dot{q})$ also varies contravariantly. These transformation properties preclude the density---in the Lagrangian formalism---from being coordinate-invariant in general. Of course, there exist special coordinate systems that allow for faithful point-wise comparison of the density. For instance, Cartesian coordinates---for a system in an inertial frame of reference---support this formal representational property. Yet, we still need to know which coordinate system we are in (or have transformed to) before making these point-wise comparisons. In general, non-Cartesian or non-inertial coordinate systems will not allow for point-wise comparisons of the density. The Hamiltonian formalism circumvents this further epistemic requirement by defining the density using quantities that vary covariantly. Hence, we can perform point-wise comparisons in the Hamiltonian formalism without knowing what coordinate system we are in.

[Should we briefly illustrate or gesture at an illustration of a failure of point-wise comparison? E.g., consider a situation where the density is a constant $k$ as a function of Cartesian coordinates (x, y). If we transform to polar coordinates $(r,\theta) $, then the scalar value of the density is no longer a constant (due to the Jacobian of this transformation not being unity). For instance, the scalar value of the density increases as we go further out, even though the density is physically the same. Hence, to determine whether there is more stuff at one location in state space vs. another, we need to know which coordinate system we are in, and how the density depends on these coordinates (i.e. we can't just compare the values of $\rho$ in these coordinates and see which is larger).]

% Connect this back with fundamentality: epistemic notion of the fundamental (don't need to know which coordinate system we are in), along with a physical sense: supporting a physical invariant, namely the invariant phase space density. Showing that this is more physically significant than the state variables themselves.

Section~\ref{fundamentality} analyzed fundamentality in terms of generality. We argued that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in part because it is more general. By considering how Hamiltonian mechanics supports faithful point-wise density comparisons, we have seen two further notions of fundamentality. The first is epistemic: by representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in. Any and all (time-independent) coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See~\ref{gauge} for discussion.} Not only is this formal property methodologically advantageous, but it also bespeaks a further physical significance. This brings us to a second, physical notion of fundamentality: the phase space density $\rho$ is physically fundamental because it tracks how much of the system is in each possible state. It quantifies the amount of stuff in each phase space state. Section~\ref{deterministic} demonstrated what is perhaps a third reason for viewing the density as fundamental: by defining the density, we recover the right mathematical structure, namely a symplectic manifold. If we did not define the density, then we would need another explanation of why this mathematical structure arises for classical systems. [I think this final point might be repetitive given a claim made in the next paragraph]
%see 11:50am of 12-21 meeting video for clarification of this point. There are probably finer details for me to incorporate.

>>>>

As indicated at the start of Section~\ref{privileging}, we use ``fundamentality" in a non-metaphysical fashion. On our usage, differences in fundamentality amount to differences in generality.\footnote{Of course, we allow for other non-metaphysical uses of ``fundamental," tracking differences in epistemic or conceptual primacy. For instance, we can think of the canonical position and momentum observables as ``epistemically fundamental" because they suffice to completely specify a classical state, fixing its values for all other observables. \textcites[31, 200]{Ruetsche} refers to this as ``fundamental in the physicist's sense." Section~\ref{density}, discusses another sense in which the Hamiltonian formalism is more fundamental than Lagrangian mechanics.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of ``structure." Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires that fewer physical conditions are satisfied. Compared with North's \parencites*[]{North} account of structure, our definition has two distinct advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which empiricists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

\begin{comment}
* ``Fundamentality" because it makes the physics more prominent. Some people use coordinate-free/independence as necessary conditions for fundemantaly/physical significance. This means coordinates do not appear in the definition. Yet, densities require coordinate for the definition, therefore cannot be coordinate-free/independent, therefore there are some physically significant quantities that do not satisfy that. Why fundamental??? (reverse mathematics? fundamental because it recovers the correct math from the physical assumption?)

@@

We have now shown that Hamiltonian mechanics requires less structure than Lagrangian mechanics because it requires fewer physical conditions. While this provides one reason for privileging Hamiltonian mechanics, it is not the only one. A special relationship exists The special relationship between Hamiltonian mechanics and invariant densities provides a further---no less important---reason for privileging Hamiltonian mechanics. In this section, we argue that viewing physics through the lens of invariant densities has much to commend it. In general, it provides a better understanding of classical systems.

In the Hamiltonian formalism, the density has the following desirable property: we can compare the density point-wise across phase space. If the scalar value of the density at one point is larger than another, this entails that the density is higher at that point of phase space. Physically, this means that there is more physical stuff at this point. Likewise, if the scalar value of the density is the same at two different points, this entails that there is the same amount of stuff at those points. For brevity, will refer to this as the property of \textit{point-wise comparison}. Point-wise comparison requires a coordinate-invariant density. Defining an invariant density requires, in turn, that we define state space over a symplectic manifold, as the derivation in Section~\ref{deterministic} shows. In other words, to define a density with the property of point-wise comparison, we must use canonical coordinates $(q, p)$ defined on a symplectic manifold.

Section~\ref{dependence} shows why point-wise comparison does not hold in general: the density is a coordinate-dependent quantity, and thus it is not generically coordinate-invariant. Consequently, within the Lagrangian formalism, point-wise comparison holds only within special coordinate systems. This gives a further reason to privilege Hamiltonian mechanics. Section~\ref{fundamental} argues for the fundamental physical significance of the phase space density. Since the density is coordinate-dependent, this shows that coordinate-independence is not a necessary condition for fundamental physical significance. Our position undercuts metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms [get citations: Field 1980; Chen papers; North]. Additionally, we explain how the conjugate momentum $p$ is itself a gauge-dependent quantity, bolstering our claim that the density is fundamental, not the state variables. Hence, the real physical significance of canonical coordinates is that they allow us to define an invariant phase space density $\rho $.

\subsection{Coordinate-dependence and coordinate-variance}
\label{dependence}

%It is helpful to take a quick look at how densities work in general so that we can better understand what is so special about them in the context in Hamiltonian mechanics.

To understand what makes densities special in the context of Hamiltonian mechanics, it is helpful to consider densities in general. By a `density' we mean the intuitive physical notion of a quantity divided by the region it occupies, like mass per volume or charge per area. Over a continuum, this is an integrable function $\rho(x)$. Mathematically, the concept is captured by the Radon–Nikodym derivative: we have (i) a region $U$, (ii) two measures $m$ and $V$, and (iii) the derivative satistifies $m(U) = \int_U \rho dV$. This relationship shows that the units of $\rho$ depend on the units of both $m$ and $V$. More precisely, if $\rho(x, y)$ is a function of multiple variables, dimensional analysis requires that the units of $\rho$ are consistent with the units of the arguments in the expression $\int_U \rho(x, y) dx dy$. In polar coordinates $(r, \theta)$, if we measure $r$ in meters and $\theta$ in radians, then $\rho(r, \theta)$ must return mass over meter-radians .

The density is coordinate dependent because its definition requires us to make a choice of coordinates. Worse than that, it is not even coordinate invariant: the density changes with the Jacobian. For example, changing from polar to Cartesian coordinates, we have $\rho(x,y) = |J| \rho(r, \theta) = r \rho(r, \theta)$. Since the Jacobian is a function of the point, the value of $|J|$ can change from point to point. Consequently, the density is not guaranteed to be directly comparable at different points: point-wise comparison need not hold in an arbitrary coordinate system. If we have a uniform, constant density in $x$ and $y$, we will have an $r$-dependence in $r$ and $\theta$. This dependence is introduced precisely because the area identified by $dr$ and $d\theta$ is $dr \,r d\theta$, which is a function of $r$. This makes the value of the density, in general, not directly comparable at different points. Hence, to determine whether there is more stuff at one location in state space than another, we need to know which coordinate system we are in, and how the density depends on these coordinates (i.e. we can't just compare the values of $\rho$ in these coordinates and see which is larger).
%The Hamiltonian formalism circumvents this further epistemic requirement by defining the density using quantities that vary covariantly. Hence, we can perform point-wise comparisons in the Hamiltonian formalism without knowing what coordinate system we are in.
%when we define the density originally, it is coordinate-free (rho as a function of the state space of the particles, S). But when we go to express it numerically, then it becomes coordinate-dependent. (I.e. when we express it over a manifold, then it needs to be a density for integration, and hence it becomes coordinate-dependent). [see 57.3 minutes of 12-21 recording for discussion]

In differential geometry, a different mathematical object is called the `density,' namely a volume form (i.e. an alternating function from $n$ vectors to a real number) that is unaffected by changes in orientation. Our physically significant density $\rho$ is actually the component of this volume form $\nu = \rho(x^i) dx^1 \wedge dx^2 \wedge ... \wedge dx^n$. Note that $\nu$ combines the density and the volume into a coordinate-free object. Considering the total mass in the region $U$, we have $m(U) = \int \nu$ so essentially $\nu = dm$ which is the infinitesimal mass contribution we are integrating. But this means that the value returned by $\nu$ depends not only on how much stuff there is at a given point, but also on the size of the volume identified by the $n$ vectors given to the form. In other words, $\nu$ has dimensions of only mass and is defined as a function of both a point and a volume element. In contrast, the density $\rho$ is defined at each point and has dimensions of mass over volume. This shows that the coordinate-independent function $\nu$ cannot play the role of the coordinate-dependent density $\rho$.

In Riemannian geometry, we have a canonical volume form $dV = \epsilon \sqrt{|g_{ab}|} dx^1 \wedge dx^2 \wedge ... \wedge dx^n$ where $|g_{ab}|$ is the determinant of the metric tensor as expressed in local coordinates, and $\epsilon$ is $1$ or $-1$ depending on the orientation. In component notation, we can express the volume form as $\sqrt{|g_{ab}|}\epsilon_{a_1 ... a_n}$, where $\epsilon_{a_1 ... a_n}$ is the Levi-Civita symbol. We can then express any differential density $\nu = \hat{\rho} dV$ as the product of the volume density times a scalar function $\hat{\rho} \equiv \rho \ \epsilon \sqrt{|g_{ab}|}$. At first glance, this expression misleadingly suggests that the density $\hat{\rho}$ is coordinate-independent: it is a scalar quantity with the correct dimensions of mass over volume. Nevertheless, $\hat{\rho}$ remains coordinate-dependent because it is a function of $|g_{ab}|$, which depends on the choice of coordinates and is, in general, a function of the point. However, we can always find a set of coordinates that, at least locally, makes $|g_{ab}|$ constant. Since the square root of the determinant $|g_{ab}|$ is the Jacobian of the transformation to those coordinates, $|g_{ab}|$ will be constant only when this Jacobian equals one. For example, in Cartesian coordinates $|g_{ab}| = 1$ and the volume form $dV = dx \wedge dy$; in polar coordinates $|g_{ab}| = r^2$ and hence the volume form $dV = r dr \wedge d\theta$. This means that $\hat{\rho}$ at best expresses the density in a particularly convenient Cartesian-like coordinate system: one in which we can compare the density point-wise.\footnote{This is similar to what happens with the concept of proper time: it is defined in the comoving frame, so it is coordinate dependent. Yet, the expression $d\tau = \frac{1}{c} \sqrt{g_{ab} dx^a dx^b}$ is coordinate invariant.} Again, the upshot here is that we cannot avoid using a coordinate-dependent density. Outside the context of Hamiltonian mechanics, the best we can do is obtain the property of point-wise comparison in a special coordinate system.
%We can also see this if we look at the units of $\hat{\rho}$, which are in terms of inverse volume as expressed in the privileged system. From 1-4 meeting: I thought that $\hat{\rho}$ has units of mass over volume.

We can now look at what happens for distributions over position and velocity, i.e. on the tangent bundle in Lagrangian mechanics. Note that the velocities transform in the same way (i.e. contravariantly) as differentials in position. Consequently, the density $\rho(q,\dot{q})$ also varies contravariantly. These transformation properties preclude the density---in the Lagrangian formalism---from being coordinate-invariant in general. In detail, $\rho(x^i, v^j) = \sqrt{|g_{ab}|} \sqrt{|g_{ab}|} \rho(\hat{x}^i, \hat{v}^j) = |g_{ab}| \rho(\hat{x}^i, \hat{v}^j)$. The determinant of the metric tensor is therefore directly the volume of the tangent bundle. The relationship to the density as expressed in phase space is therefore the following:
\begin{equation}
	\rho(q^i, p_j) = |g_{ab}| \rho(x^i, v^j).
\end{equation}
A change of coordinates will induce a contravariant change in the velocity and a covariant change in the momentum. The left side will remain invariant. The two terms on the right side will change in the opposite way so that their product is invariant. Focusing on densities, then, brings about a physically-motivated way to understand the relationship between the cotangent and the tangent bundle, where the metric tensor plays a role in defining the geometry of the second in a similar way that the symplectic form defines the geometry of the first: it defines the count of states so that densities can be properly tracked.

Overall, these considerations show that in the Lagrangian framework, there are at best special coordinate systems that allow for point-wise comparison of the density. In this context, we need to know which coordinate system we are in before making these comparisons. The Hamiltonian formalism circumvents this further epistemic requirement by defining the density using quantities that vary covariantly. Hence, we can perform point-wise comparisons in the Hamiltonian formalism without knowing what coordinate system we are in.
% Josh: why is the determinant of the metric tensor equal to the volume of the tangent bundle in this case? [Discussed on 1-4 meeting, 11:56 a.m.]. Perhaps study discussion of this in the long paper.

% Remember: the density is coordinate-invariant but not coordinate-independent; it is actually coordinate-dependent (i.e. it does not admit of something like a coordinate-free definition)
% Also remember: there are quantities that are both coordinate-variant and coordinate-dependent that are still physically significant, e.g. choices of q's and p's (?)
% Coordinate-invariant in general means that definition of the quantity involves referencing coordinates, but if you change the coordinates, then the value of the quantity remains the same. (E.g. transforms as a scalar). so densities are never coordinate-independent. defining a density requires a choice of units, so involves referencing units. 
%see 57.3 minutes of 12-21 recording: when we define the density originally, it is coordinate-free (rho as a function of the state space of the particles, S). But when we go to express it numerically, then it becomes coordinate-dependent. (I.e. when we express it over a manifold, then it needs to be a density for integration, and hence it becomes coordinate-dependent). 
% see 51 minutes of 12-21 recording: Densities (the coordinate-dependent kind) are physically significant partly because they tell you how much stuff is in each state

\subsection{Coordinate-dependent yet Fundamental}
\label{fundamental}

@@

*[Incorporate the following from section 5.2] Not all ``state variables" are physical. Hamiltonian mechanics does not really care what the q's and p's are. It is the motion assumption that tells us that q's are position, simply because they define the unit system. P is a gauge quantity (we care about the trajectory, the velocity). In this view, p is simply a ``label'' we can use to express the density over states. Therefore we do care about the physicality of the invariant distribution more than the physicality of the state variables themselves. $->$ Invariant density is really the important physical element/object in Hamiltonian mechanics.

* Units are physically significant. In fact, the ability to express the densities in units is what gives us the symplectic structure. It's a physical requirement that we must be able to express the density in some unit.
* Invariant densities: physical significance connected to coordinate invariance. [incorporated]
* Density comparison is independent of coordinates. We can compare densities point-wise without worry. [incorporated]

* Makes the fundamental role that phase space plays in statistical mechanics more physically salient/perspicuous. Areas/volumes are measures of number of states. They are coordinate invariant. Also the Shannon/Gibbs entropy are coordinate invariant.

% Section~\ref{action} uses the invariant density to provide a geometrical interpretation of the action principle.

% Goal of the following paragraph is to explain the various senses in which the invariant phase space density is fundamental

Section~\ref{fundamentality} analyzed fundamentality in terms of generality. We argued that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in part because it is more general. By considering how Hamiltonian mechanics supports point-wise density comparisons, we have encountered two further notions of fundamentality. The first is epistemic: by representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in. Any and all (time-independent) coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See~\ref{gauge} for discussion.} Not only is this formal property methodologically advantageous, but it also bespeaks a further physical significance. This brings us to a second, physical notion of fundamentality: the phase space density $\rho$ is physically fundamental because it tracks how much of the system is in each possible state. It quantifies the amount of stuff in each phase space state. Section~\ref{deterministic} demonstrated what is perhaps a third reason for viewing the density as fundamental: by defining the density, we recover the right mathematical structure, namely a symplectic manifold. If we did not define the density, then we would need another explanation of why this mathematical structure arises for classical systems. 
%see 11:50am of 12-21 meeting video for clarification of this point. There are probably finer details for me to incorporate.

Collectively, these considerations speak in favor of viewing the invariant phase space density as not only physically significant but also fundamental. Yet, as Section~\ref{dependence} explained, the density is always coordinate-dependent, even when coordinate-invariant. In other words, there is no coordinate-independent (or coordinate-free definition) of the invariant phase space density. Affording the density fundamental significance therefore places us at odds with those who take coordinate-independence as a necessary condition for fundamentality. [To do: provide literature references for this and motivating quotations]

Some philosophers have instead privileged the state variables $(q, p)$. For instance, in her argument for the fundamentality of Hamiltonian mechanics, North claims that ``momentum is a truly fundamental quantity, on a par with position" \parencites*[77]{North}. [To do: see if Wallace 2003 also makes this claim about the fundamentality of conjugate momentum, or the similarly flawed claim that conjugate momentum and position are on equal footing within Hamiltonian mechanics]. Yet, as we show below, the conjugate momentum $p$ is actually a gauge-redundant quantity. Assuming that gauge-invariance is a necessary condition for fundamental physical significance, this precludes $p$ from having fundamental status. Instead, we should understand the conjugate momentum $p$ as a label that we can use to express the invariant density over states. In this way, the invariant density is more physical than the state variables themselves. 
% Possibly a stronger claim: This entails that the invariant density is really the important physical element/object in Hamiltonian mechanics.

[To do: explain why we have grounds to privilege the invariant density over not only the conjugate momenta but also the canonical position coordinate. Is this connected with the following thought: Hamiltonian mechanics does not really care what the q's and p's are. It is the motion assumption that tells us that q's are position, simply because they define the unit system. ]

[To do: think about whether treating gauge-invariance as a necessary condition for fundamentality is in tension with our position, specifically our contention that coordinate-dependency does NOT preclude a quantity from fundamental physical significance. Does our position require an important physical difference between gauge-redundancy and coordinate-dependence?].
% % Note that if coherent, our position has the following consequence: gauge-redundancy is a sufficient condition for a quantity to be non-fundamental. But coordinate-dependency is NOT a sufficient condition for being non-fundamental. This could strike some readers as in tension. If we are comfortable viewing gauge-redundancy as precluding a quantity from fundamental significance, then why not coordinate-dependence?

We now explain why the conjugate momentum $p$ is gauge-dependent. Recall that motion equivalence entails the following: knowledge of the trajectory in physical space suffices for knowledge of the state in phase space. In other words, we can completely characterize states in terms of the canonical position coordinate $q$ and its first derivative $\dot{q}$. Hence, any transformation that leaves $(q,\dot{q}) $ invariant \textit{ipso facto} leaves the physical state invariant. We want such transformations to be canonical in order to preserve the invariant density. This means that the new state variables $(q, \hat{p})$ will respect the usual Poisson brackets:
\begin{equation}
\begin{aligned}
	\{q^i, q^j\} &= 0 \\
	\{q^i, \hat{p}_j\} &= \delta^i_j \\
	\{\hat{p}_i, \hat{p}_j\} &= 0
\end{aligned}
\end{equation}
Using the second equation we find:
\begin{equation}
\begin{aligned}
	\{q^i, \hat{p}_j\} &= \frac{\partial q^i}{\partial q^k} \frac{\partial \hat{p}_j}{\partial p_k} - \frac{\partial q^i}{\partial p_k} \frac{\partial \hat{p}_j}{\partial q^k} \\
	&= \delta^i_k \frac{\partial \hat{p}_j}{\partial p_k} - 0 \frac{\partial \hat{p}_j}{\partial q^k} = \frac{\partial \hat{p}_j}{\partial p_i} = \delta^i_j
\end{aligned}
\end{equation}
Integrating we have:
\begin{equation}
\begin{aligned}
\hat{p}_j = p_j + A_j(q^i)
\end{aligned}
\end{equation}
where $A_j$ are arbitrary functions of position. Using the third Poisson bracket, we have:
\begin{equation}
\begin{aligned}
0 = \{\hat{p}_i, \hat{p}_j\} &= \{p_i + A_i, p_j + A_j\} \\
&= \{p_i, p_j \} + \{p_i , A_j\} + \{A_i, p_j \} + \{A_i, A_j\} \\
&= 0 - \{A_j, p_i\} + \{A_i, p_j \} + 0 \\
&= - \frac{\partial A_j}{\partial q^i} + \frac{\partial A_i}{\partial q^j} \\
&= curl(A_j) \\
\end{aligned}
\end{equation}
Since $A_j$ is curl free, we can express it as the gradient of a function $f$. Therefore changes of state variable of the form $\hat{p}_j = p_j + \partial_j f$ do not change either the equations of motion or the trajectories. We recognize these as a family of gauge transformations that change the conjugate momentum variable $p$ while leaving $(q,\dot{q}) $ invariant. The gauge-redundancy of the conjugate momentum supplies further grounds for privileging the invariant phase space density. Additionally, it shows that not all state variables are on equal physical footing. Although the canonical position and momentum variables are on equal syntactical footing within symplectic mechanics, this does not entail that they are equally physically significant. Indeed, as we now argue, some coordinate choices are more significant than others.
% Whereas symplectic geometry creates a mathematical, syntactic symmetry between the $q$ and $p$ variables, this is not a physically significant symmetry from the standpoint of measurement.
%cite van Fraassen on the measurement point; e.g. APA session on Halvorson book

In privileging the phase space density, we have relied on its coordinate-invariance. Plausibly, coordinate-invariance is a necessary condition for fundamental physical significance. Being necessary for fundamental significance does not, however, make coordinate-invariance necessary for physical significance simpliciter. Our position here again places us at odds with others in the literature. For instance, North claims that coordinate choices are ``merely arbitrary choices of description" \parencites*[61]{North}. Yet, some choices of the coordinates $q$ are not appropriately connected with physical measurement, making them less physically meaningful [to do: provide a clear example of this]. Crucially, the coordinates $q$ define the units of physical measurement, allowing us to assign numbers to physical quantities. These units are physically significant despite not being coordinate-invariant. Consequently, although symplectic mechanics allows arbitrary mixings of $q$'s and $p$'s that preserve the symplectic manifold, not all of these combinations are physically significant. In general, the canonical position coordinates $q$ are epistemically privileged because they are linked with measurements, in a sense that the logically-derivative  $p$'s are not.

[Possible further point to include: discuss how Cartesian coordinates can be physically significant, in virtue of position and velocity being independent degrees of freedom, whereas this independence does not hold generically across coordinate systems. (In contrast, position and conjugate momentum variables always characterize separable/independent degrees of freedom).]

[Not sure if/where to include the following point (possibly with the discussion of the coordinate-dependence of the density?): As the derivation in Section~\ref{deterministic} shows, a necessary condition for defining the symplectic structure is that we can express the density in a particular choice of units. If we did not have the physical ability to define the density in some units we would not be able to define this structure. ]

[To do: expand on the following point: ] This is why statistical mechanics works best on phase space (we have a uniform measure over q/p). Focusing on the density makes the fundamental role that phase space plays in statistical mechanics more physically salient/perspicuous. Areas/volumes are measures of number of states. They are coordinate invariant. Also the Shannon/Gibbs entropy are coordinate invariant.
